---
title: "3. Running the inference"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{3. Running the inference}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Running the inference

Once all inputs are in place and the TOSCA object has been correctly initialised, the inference is launched with a single call to `fit()`. The function takes the TOSCA object as input and populates its inference results slot. For resistances mediated by a CNA event, as is the case for UPN06, `model_name` should be set to `"CNA"`. The `dormancy` argument should be set to `TRUE` when the patient has received a chemotherapeutic agent known to induce tumour cell dormancy, such as cytarabine. Additional arguments allow fine-grained control over the MCMC sampler (number of chains, iterations per chain, warmup length, and so on); see `?TOSCA::fit` for full details.

```{r, echo=FALSE, message=FALSE, warning=FALSE}
library(TOSCA)
set.seed(10)
data("UPN06")
```

```{r, message=FALSE, warning=FALSE, eval=FALSE}
UPN06 <- TOSCA::fit(
  UPN06,
  model_name = "CNA",
  dormancy   = TRUE
)
```


Upon completion, `fit()` prints a summary report containing:

- **Convergence diagnostics:** whether all $\hat{R} < 1.01$, the number of divergent transitions, and whether the E-BFMI fell below 0.3 in any chain.
- **Preliminary posterior predictive check:** whether the observed mutation count for each group falls between the mode and the 90th percentile of the posterior predictive distribution.
- **Parameter summary:** posterior estimates for each inferred parameter, including the posterior predictives.

A summary of the inference results can be retrieved at any time with:

```{r, message=FALSE, warning=FALSE}
TOSCA::get_fit_summary(UPN06)
```

The fitted object also exposes the raw posterior draws, model specification, and diagnostic summary directly:

```{r, message=FALSE, warning=FALSE}
UPN06$Fit$posteriors
UPN06$Fit$model_info
UPN06$Fit$diagnostic_summary
```

