---
title: "4. Visualising the results"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{3. Running the inference}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

### Timing posteriors

The inferred timing posteriors can be visualised on the patient's clinical timeline in two complementary ways: as histograms overlaid on a single timeline, or as stacked densities. In both plots, treatment periods are shown as coloured rectangles and sampling dates as points on the timeline.

```{r, echo = FALSE, message=FALSE, warning=FALSE}
library(TOSCA)
set.seed(10)
data("UPN06")

# UPN06 <- TOSCA::fit(
#   UPN06,
#   model_name = "CNA",
#   dormancy   = TRUE
# )
```

```{r, message=FALSE, warning=FALSE, fig.width=7, fig.height=4, dpi = 100}
TOSCA::plot_timing(UPN06)
```

```{r, message=FALSE, warning=FALSE, fig.width=8, fig.height=5, dpi = 100}
TOSCA::plot_timing_MAP(UPN06)
```

By default the x-axis is expressed in calendar dates. If preferred, timings can be displayed in days relative to a clinically meaningful reference event. For UPN06, we use the allo-HSCT date as the origin:

```{r, message=FALSE, warning=FALSE, fig.width=7, fig.height=4, dpi = 100}
allo_HSCT <- "2014-04-23"
TOSCA::plot_timing_days(UPN06, days_from = allo_HSCT, event_name = "allo-HSCT")
```

Summary statistics can likewise be computed relative to a reference date using `days_from()`. This is particularly useful for classifying the timed event as *de novo* (originating after the reference) or *pre-existing* (originating before it). For CNA-mediated resistance, the parameter `t_cna` and the length of the corresponding CNA, as specified in the mutations input, must be provided:

```{r, message=FALSE, warning=FALSE}
TOSCA::days_from(UPN06, from = allo_HSCT, parameter = "t_cna", cna_length = 47911693)
```

Consistent with the findings reported in the reference paper, the posterior places the CNLOH of the HLA locus after the allo-HSCT and in close temporal proximity to ganciclovir exposure, supporting a *de novo* origin of the resistance event.

### Dynamic parameters

To assess the inferred dynamic parameters, `plot_prior_vs_posterior()` displays the prior density alongside the posterior histogram for each parameter. The posterior median is indicated by a vertical red line.

```{r, message=FALSE, warning=FALSE, fig.width=8, fig.height=4, dpi = 100}
TOSCA::plot_prior_vs_posterior(UPN06)
```

### Posterior predictive check

To verify that the inferred parameters are compatible with the observed data, `plot_ppc()` displays the multivariate posterior predictive distribution (PPD) for each mutation group. The observed mutation count is shown as a red point; the PPD is represented as a cloud of draws, colour-coded by their distance from the observed value.

```{r, message=FALSE, warning=FALSE, fig.width=5, fig.height=5, dpi = 100}
TOSCA::plot_ppc(UPN06)
```

## MCMC diagnostics

TOSCA exposes a suite of diagnostic plots built on the [bayesplot](https://mc-stan.org/bayesplot/) package. These should be inspected routinely to verify that the sampler has explored the posterior adequately. Available diagnostics include chain trace plots, $\hat{R}$ plots, effective sample size (ESS) plots, and divergent transitions.

```{r, message=FALSE, warning=FALSE, fig.width=7, fig.height=7, dpi = 100, eval = FALSE}
TOSCA::plot_mcmc_chains(UPN06)
```

```{r, message=FALSE, warning=FALSE, fig.width=3, fig.height=3, dpi = 100, eval = FALSE}
TOSCA::rhats_plot(UPN06)
```

```{r, message=FALSE, warning=FALSE, fig.width=3, fig.height=3, dpi = 100, eval = FALSE}
TOSCA::ess_plot(UPN06)
```

```{r, message=FALSE, warning=FALSE, fig.width=7, fig.height=3, dpi = 100, eval = FALSE}
TOSCA::plot_divergent_transitions(UPN06)
```

